---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a00_introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is Syrona?

**Syrona** compares health datasets built on the OMOP Common Data Model. Given two OMOP CDM databases (or two cohorts within the same database), it:

1.  **Extracts** stratified prevalence tables across three clinical domains (conditions, procedures, drugs)
2.  **Compares** paired datasets by computing log2 prevalence ratios per stratum
3.  **Synthesizes** (pools) results via multi-level random-effects meta-analysis

The output is a set of CSV tables that can be explored in the Syrona dashboard or consumed by downstream tools.

## How it differs from other OHDSI tools

| Tool | Purpose |
|-----------------------------|-------------------------------------------|
| ACHILLES | Profile a single database (aggregate statistics) |
| CohortDiagnostics | Validate cohort definitions (incidence, attrition) |
| CohortContrast | Feature selection: target vs control within one database |
| **Syrona** | **Compare the same population across different databases/sites** |

Syrona is designed for multi-site comparisons where you want to understand how prevalence patterns differ between institutions, countries, or data sources.

## Installation

```{r, eval=FALSE}
# From GitHub:
# install.packages("remotes")
remotes::install_github("HealthInformaticsUT/syrona")
```

### Dependencies

-   [CDMConnector](https://CRAN.R-project.org/package=CDMConnector) (\>= 2.0.0) for database access
-   [omopgenerics](https://CRAN.R-project.org/package=omopgenerics) (\>= 1.3.0) for cohort table operations
-   [meta](https://CRAN.R-project.org/package=meta) for random-effects meta-analysis
-   [duckdb](https://CRAN.R-project.org/package=duckdb) for local DuckDB files

## Quick start

### 1. Connect to a database

```{r, eval=FALSE}
library(syrona)

# Local DuckDB file (read-only by default)
db <- syrona_connect("path/to/omop.duckdb")

# PostgreSQL (e.g. via SSH tunnel)
db <- syrona_connect_pg(
  dbname = "omop",
  user = "analyst",
  cdm_schema = "cdm",
  write_schema = "results_analyst"
)
```

### 2. Extract a dataset

```{r, eval=FALSE}
# Extract all three domains
tables <- extract_all("Dataset_A", db = db)

# Or a single domain (faster for large databases)
tables <- extract_all("Dataset_A", db = db, domains = "conditions")
```

### 3. Compare two datasets

```{r, eval=FALSE}
# After extracting both datasets:
compare_all("Dataset_A", "Dataset_B")
```

### 4. Disconnect

```{r, eval=FALSE}
syrona_disconnect(db)
```

## Output structure

Syrona writes CSV files to two directories:

```         
data/
  sources/                        # Phase 1: extracted datasets
    Dataset_A/
      _metadata.csv
      condition_prevalence.csv    # concept x year x sex x age_group
      condition_info.csv          # concept metadata
      condition_chapters.csv      # SNOMED/ICD-10 chapter assignments
      condition_attributes.csv    # SNOMED relationship targets
      demographics.csv            # birth year x sex counts
      death_counts.csv            # deaths by stratum
      ...                         # same pattern for procedures + drugs
    Dataset_B/
      ...
  comparisons/                    # Phase 2-3: comparison results
    Dataset_A_vs_Dataset_B/
      _metadata.csv
      condition_yearly.csv        # per-stratum prevalence ratios
      condition_meta_agegroups.csv # meta across years
      condition_meta_by_sex.csv   # meta across age groups
      condition_meta_summary.csv  # final summary (one row per concept)
      ...                         # same pattern for procedures + drugs
```

## Next steps

-   [Extraction](a01_extraction.html) - detailed extraction guide with all parameters
-   [Comparison](a02_comparison.html) - comparison pipeline and meta-analysis
-   [Cohorts](a03_cohorts.html) - OHDSI cohort-based extraction workflows
