---
title: "Comparison & Meta-Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a02_comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Phases 2-3 compare two extracted datasets by computing stratified prevalence ratios and synthesizing them via multi-level random-effects meta-analysis.

The pipeline has 4 steps, each producing a more aggregated table:

```         
yearly               concept x year x sex x age_group    (raw PR per stratum)
  |
  v  meta across years
meta_agegroups       concept x sex x age_group           (pooled across years)
  |
  v  meta across age groups
meta_by_sex          concept x sex                       (pooled across age groups)
  |
  v  meta across sexes
meta_summary         concept (sex = "Both")              (final summary)
```

## Running a comparison

### All domains

```{r, eval=FALSE}
library(syrona)

# Both datasets must already be extracted (in data/sources/)
compare_all("Hospital_A", "Hospital_B")
```

This loads both datasets, runs the 4-step pipeline for each domain present in both, and saves results to `data/comparisons/Hospital_A_vs_Hospital_B/`.

### Single domain

```{r, eval=FALSE}
compare_all("Hospital_A", "Hospital_B", domains = "conditions")
```

### Without saving

```{r, eval=FALSE}
results <- compare_all("Hospital_A", "Hospital_B", save = FALSE)
names(results)
#> [1] "condition_yearly" "condition_meta_agegroups"
#> [3] "condition_meta_by_sex" "condition_meta_summary"
```

## Step-by-step usage

You can also run each step individually for more control:

```{r, eval=FALSE}
d1 <- load_dataset("Hospital_A")
d2 <- load_dataset("Hospital_B")

# Step 1: Yearly comparison (inner join on concept x year x sex x age_group)
yearly <- compare_yearly(d1, d2, prev_table = "condition_prevalence")

# Step 2: Meta across years
meta_ag <- compare_meta_agegroups(yearly)

# Step 3: Meta across age groups
meta_sex <- compare_meta_by_sex(meta_ag)

# Step 4: Meta across sexes
meta_sum <- compare_meta_summary(meta_sex)
```

## Prevalence ratio formula

For each matched stratum, Syrona computes:

**Log2 prevalence ratio:**

```         
log2_pr = log2(p2 / p1)
```

Where p1 and p2 are the prevalence values in dataset 1 (reference) and dataset 2 (comparison).

**Standard error (on log2 scale):**

```         
SE = sqrt((1 - p1) / (p1 * n1) + (1 - p2) / (p2 * n2)) / ln(2)
```

Where n1 and n2 are the denominators (persons observed in that stratum).

**95% confidence interval:**

```         
CI = log2_pr +/- 1.96 * SE
```

**Significance:** A stratum is significant if the CI excludes zero on the log2 scale (i.e., the fold difference excludes 1.0).

## Output columns

Each comparison table contains these columns:

| Column | Description |
|----|----|
| `concept_id` | OMOP concept ID |
| `log2_pr` | Log2 prevalence ratio |
| `se` | Standard error (log2 scale) |
| `ci_low`, `ci_high` | 95% CI bounds (log2 scale) |
| `fold_diff` | Natural-scale fold difference (2\^log2_pr) |
| `fold_ci_low`, `fold_ci_high` | Natural-scale CI bounds |
| `fold_symmetric` | max(fold, 1/fold) - useful for ranking |
| `p_value` | Two-sided p-value |
| `sig` | TRUE if CI excludes zero |
| `meta_model_type` | "Random-Effects", "Fixed-Effect", or "Pass-Through" |
| `n_strata` | Number of strata pooled |
| `tau2` | Between-study variance (heterogeneity) |
| `I2` | Percentage of variance due to heterogeneity |
| `Q` | Cochran's Q statistic |
| `pval_Q` | P-value for Q test |

## Meta-analysis strategy

For each group of strata to be pooled, Syrona uses this fallback strategy:

1.  **Single stratum** - Pass-Through (returns the estimate as-is, no pooling)
2.  **Multiple strata** - Random-effects meta-analysis with Paule-Mandel tau estimator (`meta::metagen`)
3.  **If RE fails** - Fixed-effect (common-effect) model
4.  **If both fail** - Group is skipped

The Paule-Mandel estimator is preferred because it handles small numbers of studies better than DerSimonian-Laird.

## Heterogeneity metrics

The meta-analysis cascade produces heterogeneity metrics at each aggregation level:

-   **tau2** - between-study variance. High tau2 means the prevalence ratio varies substantially across strata (e.g., a condition that is more prevalent in young women but less prevalent in older men).
-   **I2** - percentage of total variance attributable to heterogeneity (0-100%). I2 \> 75% is conventionally considered "high".
-   **Q** and **pval_Q** - Cochran's Q test for heterogeneity. A significant Q (p \< 0.05) means the strata are not homogeneous.
-   **lower_predict**, **upper_predict** - prediction interval (where a new study's effect would likely fall).

These metrics help identify concepts where the summary prevalence ratio masks meaningful variation across strata.

## Loading and listing comparisons

```{r, eval=FALSE}
# List available comparisons
list_comparisons()
#> [1] "Hospital_A_vs_Hospital_B"

# Load a comparison
comp <- load_comparison("Hospital_A", "Hospital_B")
names(comp)
#> [1] "condition_yearly" "condition_meta_agegroups"
#> [3] "condition_meta_by_sex" "condition_meta_summary"
```
