---
title: "OHDSI Cohort Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a03_cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

By default, `extract_all()` extracts data for all persons in the database. But often you want to compare specific subpopulations - patients from a particular hospital, a clinical trial cohort, or a phenotype definition.

Syrona supports OHDSI-standard cohort tables for this. A cohort table has 4 columns:

| Column | Type | Description |
|--------|------|-------------|
| `cohort_definition_id` | INTEGER | Which cohort |
| `subject_id` | INTEGER | person_id |
| `cohort_start_date` | DATE | When they enter |
| `cohort_end_date` | DATE | When they exit |

When you pass a `cohort_id` to `extract_all()`, it restricts extraction to cohort members and clips all events to each person's cohort window.

## Care-site cohorts

The most common use case: compare hospitals within the same OMOP CDM database.

```{r, eval=FALSE}
library(syrona)

# Connect with write access (needed to create cohort table)
db <- syrona_connect_pg(
  dbname = "omop",
  user = "analyst",
  cdm_schema = "ohdsi_cdm_202511",
  write_schema = "results_analyst"
)

# Find available care sites
list_care_sites(db$con, cdm_schema = "ohdsi_cdm_202511")
#>   care_site_id    care_site_name  n_patients
#> 1          101   Central Hospital      45000
#> 2          205   University Clinic     28000
#> 3          312   Regional Hospital     15000
```

### Create cohorts

```{r, eval=FALSE}
# Create a cohort for each hospital
create_caresite_cohort(
  con = db$con,
  care_site_id = 101,
  cohort_id = 1,
  cohort_schema = "results_analyst",
  cdm_schema = "ohdsi_cdm_202511"
)

create_caresite_cohort(
  con = db$con,
  care_site_id = 205,
  cohort_id = 2,
  cohort_schema = "results_analyst",
  cdm_schema = "ohdsi_cdm_202511"
)
```

Each person's cohort window runs from their first visit start date to their last visit end date at that care site. By default, these are clipped to observation period overlap (`restrict_to_observation = TRUE`).

### Extract by cohort

```{r, eval=FALSE}
# Extract each hospital's data
extract_all("Central_Hospital", db = db,
            cohort_id = 1, cohort_schema = "results_analyst")

extract_all("University_Clinic", db = db,
            cohort_id = 2, cohort_schema = "results_analyst")

# Then compare
compare_all("Central_Hospital", "University_Clinic")
```

### Verify a cohort

```{r, eval=FALSE}
cohort_summary(db$con, cohort_id = 1, cohort_schema = "results_analyst")
#>   cohort_definition_id n_entries n_persons  min_start    max_end
#> 1                    1     45000     45000 2005-01-03 2023-11-28
```

### Delete a cohort

```{r, eval=FALSE}
delete_cohort(db$con, cohort_id = 1, cohort_schema = "results_analyst")
```

## Cohorts from data frames

If you already have cohort membership computed locally (from a CSV, a phenotype algorithm, or another tool), use `insert_cohort()` to upload it via the modern omopgenerics pathway:

```{r, eval=FALSE}
# Build a cohort data frame
cohort_df <- data.frame(
  cohort_definition_id = 1L,
  subject_id = c(1001L, 1002L, 1003L),
  cohort_start_date = as.Date("2015-01-01"),
  cohort_end_date = as.Date("2020-12-31")
)

# Upload to the CDM (requires writeSchema)
db$cdm <- insert_cohort(db$cdm, cohort_df, name = "my_cohort")

# The cohort is now a table in the database, usable by Syrona and other OHDSI tools
```

Note: `insert_cohort()` validates that cohort dates fall within each person's observation period, as required by the OHDSI standard.

## Care-site cohort options

`create_caresite_cohort()` has two optional parameters for fine-tuning:

### Collapse strategy

```{r, eval=FALSE}
# Default: one row per person (first visit start to last visit end)
create_caresite_cohort(..., collapse_strategy = "person_span")

# Alternative: one row per visit (multiple cohort entries per person)
create_caresite_cohort(..., collapse_strategy = "visit_occurrence")
```

`person_span` is the default and recommended for Syrona extraction - it produces one cohort window per person, which is the standard expectation.

`visit_occurrence` keeps each visit as a separate entry. This can be useful for visit-level analyses but means a person may have multiple overlapping windows.

### Observation period clipping

```{r, eval=FALSE}
# Default: clip to observation period overlap
create_caresite_cohort(..., restrict_to_observation = TRUE)

# Skip clipping (use raw visit dates)
create_caresite_cohort(..., restrict_to_observation = FALSE)
```

Clipping ensures that the cohort window doesn't extend beyond the person's known observation period. This prevents counting events that happened outside the data capture window.

## How cohort filtering works

When `extract_all()` receives a `cohort_id`, it calls `apply_cohort_filter()` which modifies the CDM table references:

- **person**: filtered to cohort members only
- **observation_period**: clipped to the intersection of observation period and cohort window (affects the ACHILLES-116 denominator)
- **death**: restricted to deaths within the cohort window
- **condition_occurrence, procedure_occurrence, drug_exposure**: events restricted to within the cohort window

This means all downstream extraction functions (denominators, prevalence, chapters, attributes) automatically operate on the filtered population without any changes to their logic.

## Compatibility with other OHDSI tools

The cohort table created by Syrona follows the standard OHDSI schema, so it is compatible with:

- **CohortDiagnostics** - for cohort validation and diagnostics
- **FeatureExtraction** - for baseline characterization
- **CohortConstructor** - for matching and cohort operations

```{r, eval=FALSE}
# Example: use CohortConstructor to create a matched control cohort
library(CohortConstructor)
db$cdm$matched <- matchCohorts(db$cdm$my_cohort, ratio = 2, name = "matched")
```
